---
title: "Bitcoin Price Prediction: A Comparison of the Predictive Accuracy of Forecasting and Machine Learning Techniques"
author:
  - name: "Bach Phung ^[Email: phungbq@miamioh.edu]"
    affiliation: Farmer School of Business, Miami University
  - name: "Lin Lu ^[Email: llu@fairfield.edu | Website: <a href=\"https://facultyprofile.fairfield.edu/?uname=llu\">Fairfield University Official</a>]"
    affiliation: Charles F. Dolan School of Business, Fairfield University
  - name: "Waldyn Martinez ^[Email: martinwg@miamioh.edu | Phone: +1-513-529-2154 | Website: <a href=\"https://miamioh.edu/fsb/directory/?up=/directory/martinwg\">Miami University Official</a>]"
    affiliation: Farmer School of Business, Miami University
  - name: "Fadel M. Megahed ^[Email: fmegahed@miamioh.edu | Phone: +1-513-529-4185 | Website: <a href=\"https://miamioh.edu/fsb/directory/?up=/directory/megahefm\">Miami University Official</a>]"
    affiliation: Farmer School of Business, Miami University
bibliography: refs.bib
csl: apa.csl
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    css: custom.css
    code_folding: show
    code_download: TRUE
    number_sections: TRUE
    paged_df: TRUE
    toc: TRUE
    toc_float: TRUE
    theme: readable
  includes:
    in_header: structure.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE,
                      echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      progress = FALSE, 
                      verbose = FALSE,
                      dev = c('png', 'pdf', 'tiff', 'postscript'),
                      fig.retina = 2,
                      out.width = '100%',
                      fig.asp = 0.7)

options(qwraps2_markup = "markdown")

# Setting properties for the default theme_bw() behavior for all plots
if(require(ggplot2) == FALSE) install.packages("ggplot2")
library(ggplot2) ; theme_set(theme_bw(base_size = 11) + theme(legend.position = 'top')) 

# Setting default color palettes to RColorBrewer Palettes
if(require(RColorBrewer) == FALSE) install.packages("RColorBrewer")
scale_colour_discrete = scale_color_brewer(palette = "Dark2")

# Setting the random seed and chunk dependencies
knitr::opts_chunk$set(cache.extra = set.seed(2021),
                      autodep = TRUE) 

```


# R Setup and Required Packages
In this project, the open-source R programming language is used to model the price of Bitcoin using forecasting and machine learning techniques. R is maintained by an international team of developers who make the language available at [The Comprehensive R Archive Network](https://cran.r-project.org/). Readers interested in reusing our code and reproducing our results should have R installed locally on their machines. R can be installed on a number of different operating systems (see [Windows](https://cran.r-project.org/bin/windows/), [Mac](https://cran.r-project.org/bin/macosx/), and [Linux](https://cran.r-project.org/bin/linux/) for the installation instructions for these systems). We also recommend using the RStudio interface for R. The reader can [download RStudio](http://www.rstudio.com/ide) for free by following the instructions at the link. For non-R users, we recommend the [Hands-on Programming with R](https://rstudio-education.github.io/hopr/packages.html) for a brief overview of the software's functionality. Hereafter, we assume that the reader has an introductory understanding of the R programming language.

In the code chunk below, we load the packages used to support our analysis. Note that the code of this and any of the code chunks can be hidden by clicking on the 'Hide' button to facilitate the navigation. **The reader can hide all code and/or download the Rmd file associated with this document by clicking on the Code button on the top right corner of this document.** Our input and output files can also be accessed/ downloaded from [fmegahed/bitcoin_pred](https://github.com/fmegahed/bitcoin_pred).


```{r packages, cache=FALSE}
# Check and install if these packages are not found locally on machine
# If pacman and devtools are not on the machine, install them
if(require(pacman)==FALSE) install.packages("pacman")
if(require(devtools)==FALSE) install.packages("devtools")

# Using pacman to install/load the required packages
pacman::p_load(
  # data analysis packages
  tidyverse, tidyquant, tidymodels, timetk, magrittr,
  # for time-series forecasting
  fpp2, timetk, modeltime, 
  # to extract models from tidy models output
  sweep, 
  # to read json data
  jsonlite, 
  # for nicely printed outputs
  DT, pander, stargazer, knitr,
  # for exploratory data analysis
  skimr, DataExplorer,
  # for plots
  scales, plotly) 
```

---

# Extracting the Historical Daily Data for Bitcoin

In this project, we chose to use the [CryptoCompare API](https://min-api.cryptocompare.com/documentation?key=Historical&cat=dataHistoday) to extract the daily Bitcoin ($BTC) OHLCV data. Note that the use of the API requires the creation of an `cc_api_key`, which can be created for free at <https://www.cryptocompare.com/cryptopian/api-keys>. To be able to run the code chunk below, the reader should either:  

- Replace the string stored in `cc_api_key.R` with their API key; or alternatively   
- Store their api_key in an .Rprofile file, which is what we are doing in this example to ensure the privacy of our api_keys.  

```{r extractBitcoin, results='asis'}
# Option 1: replace string with your api key and remove the # below
# cc_api_key = "Your API key here" 

# Option 2: include the cc_api_key = "Your API key here" in your .Rprofile
# we are using option 2 & using this project variable below

endingDate = ymd(Sys.Date() - 1) # our ending date of choice
print( paste('Our ending date for the bitcoin data is', endingDate) )

# The unix timestamp for our ending date is computed as follows
unix_timestamp = as.POSIXct(endingDate, format = "%Y-%m-%d", 
                            tz = "UTC") %>%   as.numeric()

secs_per_day = 60*60*24 # number of seconds per day

# Number of days between ending date and beginning of 2010
num_days = seq.Date(from = endingDate, to = ymd("2010-01-01"),
                    by = "-1 day") %>% length()

# Sequence of days from endingDate to 2010-01-01 in unix time
days_unix_time = seq(from = unix_timestamp, by = -secs_per_day, 
                     length.out = num_days)

# The 3 different API calls that are needed per the CC API
base_url = "https://min-api.cryptocompare.com/data/v2/histoday?fsym=BTC&tsym=USD&limit=1999&toTs="
urls = c(
  paste0(base_url,days_unix_time[1], "&api_key=", cc_api_key),
  paste0(base_url,days_unix_time[2001], "&api_key=", cc_api_key),
  paste0(base_url,days_unix_time[4001], "&api_key=", cc_api_key)
  )

# Making the 3 API calls and extracting the JSON data into a list
bitcoin_list = map(.x = urls, .f = fromJSON)

# A custom function to extract the returned data from the call
unlist_cc = function(x){
  temp = x$Data$Data
}

# Creating one data.frame of all the data
btc_df = map_df(.x = bitcoin_list, .f = unlist_cc)

# Creating a proper date from the unix timestamp
btc_df %<>% mutate(date = as_datetime(time) %>% as_date) %>% 
  arrange(date)

# Storing only rows where close >= 0.01
btc_df %<>% filter(close >= 0.01)

# checking to see if this will work from here
write_rds(btc_df, 
          '../results/btc_histDay.rds')
```


---

# Time-Series Approaches for Modeling Changes in BTC Price Over Time


```{r statModels}
nested_df = btc_df %>% 
  # selecting only the date and closing price
  select(date, close) %>%
  # creating splits, with a prediction goal of 1-day ahead
  # training data grows over time, with the first dataset ends
  # on Dec 31, 2018
  rolling_origin(
    initial = which(btc_df$date == "2018-12-31"),
    assess = 1,
    cumulative = TRUE) %>%
  # creating the following columns/variables in our nested data
  mutate(
    # extracting the training data from the splits
    data = map(.x = splits, .f = analysis),
    # extracting the training dates
    trainDates = map(.x = data, .f = extract2, 'date'),
    # extracting the ts of BTC closing price used for training
    trainClose =  map(.x = data, .f = extract2, 'close'),
    # extracting the 1-day ahead target data from the splits
    target_data = map(.x = splits, .f = assessment),
    # extracting the 1-day ahead date used for testing
    targetDate = map(.x = target_data, .f = extract2, 'date'),
    # extracting the 1-day ahead closing price used for testing
    targetClose =  map_dbl(.x = target_data, .f = extract2, 'close'),
    )

ts_results = nested_df %>% 
  # removing unneeded columns for time-series forecasting
  select(-c(splits, data, trainDates, target_data, targetDate))

ts_results %<>% 
  # forecasting techniques
  mutate(
    # naive (random walk) forecast
    naive = map(.x = trainClose, .f = naive, h = 1) %>% 
      map_dbl(.f = extract2, "mean"),
    # simple exponential smoothing forecast
    ses = map(.x = trainClose, .f = ses, h = 1) %>% 
      map_dbl(.f = extract2, "mean"),
    # linear exponential smoothing (holt's) forecast
    holt = map(.x = trainClose, .f = holt, h = 1) %>% 
      map_dbl(.f = extract2, "mean"),
    # auto.arima 1-day ahead forecast
    arima = map(.x = trainClose, .f = auto.arima) %>% 
      map(.f = forecast, h = 1) %>% 
      map_dbl(.f = extract2, "mean")
  )
  

write_rds(ts_results, 
          '../results/initial_4_models.rds')
```


# References {-}
<div id="refs"></div>


---


# Appendix {-}
In this appendix, we print all the R packages used in our analysis and their versions to assist with reproducing our results/analysis.

```{r sessionInfo}
pander::pander(sessionInfo(), compact = TRUE) # printing the session info
```